\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{times}
\usepackage{parskip}
\usepackage[hidelinks]{hyperref}
\pagestyle{empty}

\begin{document}

\noindent\textbf{Cover Letter}\\[2pt]
\noindent To: Guest Editors, IEEE Journal of Selected Topics in Signal Processing\\
\noindent Special Issue: \emph{Autonomous and Evolutive Optimization in Networked AI}\\
\noindent (Lead Guest Editor: Prof.\ Liang Song; with Profs.\ J.\ Liu, A.\ Dvir,
A.\ Skodras, V.\ C.\ M.\ Leung, and Q.\ Bi)\\[6pt]

\noindent Dear Guest Editors,

We are pleased to submit our manuscript, \textbf{``Interpretable Federated
Kolmogorov--Arnold Surrogates for Autonomous and Evolutive Resource Optimization
over Non-Stationary LEO Links,''} for consideration in the Special Issue on
\emph{Autonomous and Evolutive Optimization in Networked AI}.

\textbf{Scope and fit.} The Special Issue calls for online, self-evolving
mechanisms that let distributed models improve in multi-agent, time-varying
environments, with emphasis on the foundations of signal processing for networked
AI, networked AI for cognitive communications, and online drift detection and
compensation. Our paper addresses exactly this setting. We study autonomous
resource optimization over low Earth orbit (LEO) links, where the per-instance
optimum has no closed form and the operating regime drifts as satellites pass and
interference handovers occur. We cast the problem as learning the \emph{solution
map} of an energy-efficiency power-allocation problem, harden it into a
mixed-integer program, and decompose it into an inner statistical-learning task and
an outer communication-control task whose static optimum we derive in closed form.
A federated Kolmogorov--Arnold network learns the solution map online, while a
contextual-bandit controller learns, from feedback, the per-slot update sparsity
(the operating point implied by the Lagrangian dual price of the uplink budget),
unifying supervised learning with online decision making, as the call envisions.

\textbf{Contributions and findings.} Because the energy-efficiency objective is
flat near its optimum, the achieved utility regret is comparable across surrogates;
our claim is therefore not a raw performance win but that the KAN is the more
\emph{trustworthy} surrogate along four axes a black-box multilayer perceptron (MLP)
does not jointly provide. Over twenty seeds with paired significance tests: (i) it
is more sample-efficient on the solution map, about $1.8\times$ lower normalized
error at equal parameters (Wilcoxon $p<0.01$), while a linear predictor and a
model-based deep-unfolding solver are both clearly worse; (ii) it extrapolates to
unseen interference regimes roughly an order of magnitude more accurately, where the
parameter-matched MLP diverges on a third of the seeds; (iii) its additive-spline
structure yields closed-form, auditable design rules a black box cannot provide; and
(iv) a learned sparsification controller halves the uplink while remaining
competitive. Every component is isolated by one-factor ablations.

\textbf{Relation to prior work.} Existing learn-to-optimize methods for wireless
resource management are almost exclusively black-box networks (fully connected,
deep-unfolded, or graph-based) evaluated under stationary, centralized conditions.
Our work differs on three axes that matter for autonomous deployment: it is
interpretable (the learned policy reduces to closed-form rules), it is studied
under genuine non-stationarity with an online communication budget, and it
quantifies sample efficiency and extrapolation rather than only in-distribution
accuracy. To our knowledge this is among the first uses of Kolmogorov--Arnold
networks as interpretable optimization surrogates inside a networked, federated
signal-processing pipeline.

\textbf{Rigor and transparency.} In keeping with reproducible, trustworthy
research, all results are multi-seed with standard deviations and Wilcoxon
signed-rank tests, every comparison between model classes is parameter-matched and
trained under an identical budget, and the ablations toggle one component at a time
so that effects are attributable rather than correlational. Importantly, we report
transparently the cases where our method does \emph{not} help: online grid
self-evolution gives no significant gain over a well-sized static grid, and under
abundant centralized data or large channel-estimation error the advantage narrows.
The complete code, configurations, and result files that reproduce every number,
table, and figure in the paper are openly available.

\textbf{Declarations.} This manuscript is original, has not been published
previously, and is not under consideration for publication elsewhere. All authors
have read and approved the submission and declare no competing interests. The work
was supported in part by the Posts and Telecommunications Institute of Technology
(PTIT).

We believe the paper is a strong fit for the Special Issue and hope you find it
suitable for peer review. Thank you for your time and consideration.

\vspace{8pt}
\noindent Sincerely,\\[4pt]
\noindent Phuc Hao Do (corresponding author)\\
\noindent Center for Advanced Innovation, Research and Application,\\
\noindent Da Nang Architecture University, Da Nang, Vietnam\\
\noindent E-mail: haodp@dau.edu.vn\\[4pt]
\noindent on behalf of co-authors Huu Phu Le and Truong Duy Dinh

\end{document}
