\documentclass[journal]{IEEEtran}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{amsthm}
\newtheorem{proposition}{Proposition}
\usepackage{graphicx}
\usepackage{url}
\usepackage{booktabs}
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}
\usepgfplotslibrary{fillbetween}
\usetikzlibrary{positioning,calc}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage[hidelinks]{hyperref}

\newcommand{\pstar}{p^{\star}}

% ===================================================================
% RESULT MACROS --- PROVISIONAL (10-seed/N=10). REFRESH from the scaled
% GPU run: read results/opt.csv + paper/data/{gen,pareto}.dat and update
% every value below. All tables/prose use these macros, so editing here
% propagates everywhere. \tbc marks any value still to confirm.
% ===================================================================
\newcommand{\tbc}[1]{#1}                 % wrap placeholder values (no-op; flip to \textcolor to highlight)
% main results (Table I): NMSE, regret, uplink Mb, rule R2, params
\newcommand{\FOnmse}{\tbc{0.079}}\newcommand{\FOnmsesd}{\tbc{0.028}}
\newcommand{\FOreg}{\tbc{0.038}}\newcommand{\FObits}{\tbc{4.8}}
\newcommand{\FOrtwo}{\tbc{0.81}}\newcommand{\FOpar}{\tbc{360}}
\newcommand{\KFnmse}{\tbc{0.058}}\newcommand{\KFnmsesd}{\tbc{0.021}}
\newcommand{\KFreg}{\tbc{0.022}}\newcommand{\KFbits}{\tbc{9.0}}
\newcommand{\KFrtwo}{\tbc{0.77}}\newcommand{\KFpar}{\tbc{360}}
\newcommand{\KEnmse}{\tbc{0.080}}\newcommand{\KEreg}{\tbc{0.038}}
\newcommand{\KEbits}{\tbc{5.2}}\newcommand{\KErtwo}{\tbc{0.81}}\newcommand{\KEpar}{\tbc{404}}
\newcommand{\MLnmse}{\tbc{0.135}}\newcommand{\MLnmsesd}{\tbc{0.055}}
\newcommand{\MLreg}{\tbc{0.038}}\newcommand{\MLbits}{\tbc{9.0}}\newcommand{\MLpar}{\tbc{361}}
\newcommand{\MPnmse}{\tbc{0.170}}\newcommand{\MPnmsesd}{\tbc{0.113}}
\newcommand{\MPreg}{\tbc{0.038}}\newcommand{\MPbits}{\tbc{9.0}}\newcommand{\MPpar}{\tbc{361}}
\newcommand{\LInmse}{\tbc{2.29}}\newcommand{\LIreg}{\tbc{0.35}}
\newcommand{\LIbits}{\tbc{0.1}}\newcommand{\LIpar}{\tbc{4}}
% ratios + Wilcoxon p-values (in-text)
\newcommand{\ratioNmse}{\tbc{2.3}}\newcommand{\ratioReg}{\tbc{1.7}}
\newcommand{\pNmse}{\tbc{0.004}}\newcommand{\pReg}{\tbc{0.002}}
\newcommand{\pBits}{\tbc{0.002}}\newcommand{\pFOmlp}{\tbc{0.027}}
\newcommand{\pEvolve}{\tbc{0.014}}\newcommand{\winNmse}{\tbc{9/10}}\newcommand{\winReg}{\tbc{10/10}}
% generalization (Table II)
\newcommand{\genKANin}{\tbc{0.003}}\newcommand{\genKANex}{\tbc{0.37}}
\newcommand{\genMLPin}{\tbc{0.005}}\newcommand{\genMLPex}{\tbc{2.55}}
% extracted-rule forms + per-feature fidelity (kan_full)
\newcommand{\ruleGform}{\tbc{nonlinear}}\newcommand{\ruleGrtwo}{\tbc{0.36}}
\newcommand{\ruleIform}{\tbc{near-linear}}\newcommand{\ruleIrtwo}{\tbc{0.88}}
\newcommand{\ruleProform}{\tbc{linear}}\newcommand{\rulePrortwo}{\tbc{0.99}}
\newcommand{\ruleFid}{\tbc{0.74}}
% pareto knee + experiment scale
\newcommand{\paretoKnee}{\tbc{0.5}}
\newcommand{\Nnodes}{\tbc{24}}\newcommand{\Nseeds}{\tbc{20}}
\newcommand{\Npasses}{\tbc{5}}\newcommand{\Nsamp}{\tbc{160}}

\begin{document}

\title{Interpretable Federated Kolmogorov--Arnold Surrogates for Autonomous and
Evolutive Resource Optimization over Non-Stationary LEO Links}

\author{Phuc Hao Do$^{1}$ and Truong Duy Dinh$^{2}$
\thanks{$^{1}$Faculty of Information Technology, Da Nang Architecture University,
Da Nang, Vietnam (e-mail: haodp@dau.edu.vn).}
\thanks{$^{2}$Faculty of Information Security, Posts and Telecommunications
Institute of Technology, Ha Noi, Vietnam (e-mail: duydt@ptit.edu.vn).}
\thanks{Corresponding author: Phuc Hao Do.}}

\maketitle

\begin{abstract}
Low Earth orbit (LEO) networks must repeatedly solve resource-allocation problems
whose optima have no closed form and that drift as satellites pass and
interference regimes switch. We study an autonomous and evolutive approach in
which a federated Kolmogorov--Arnold network (KAN) learns the \emph{solution map}
of an energy-efficiency transmit-power optimization, mapping a link state to its
optimal power online. On a faithful oracle-labeled problem, a compact KAN
surrogate attains a normalized prediction error and an energy-efficiency regret
that are, respectively, about $\ratioNmse\times$ and $\ratioReg\times$ lower than a
parameter-matched multilayer perceptron (MLP) at equal parameter budget
(Wilcoxon $p<0.01$), while a linear predictor fails entirely, confirming the map
is strongly nonlinear. Crucially, the additive spline structure of the KAN lets
us \emph{read off closed-form design rules} from the trained model, a transparency
an MLP cannot provide. A contextual-bandit controller sparsifies the federated
updates, halving the uplink volume while retaining the accuracy advantage over the
MLP, tracing a favorable communication--accuracy frontier. We further report a
transparent negative result: growing the spline grid online (self-evolution) does
not improve over a well-sized static grid in this federated streaming regime. The
study positions interpretable KAN surrogates as a practical building block for
autonomous optimization in non-stationary networked-AI systems.
\end{abstract}

\begin{IEEEkeywords}
Kolmogorov--Arnold networks, federated learning, learning to optimize, LEO
satellite networks, energy efficiency, interpretability, non-stationary
environments.
\end{IEEEkeywords}

\section{Introduction}
\IEEEPARstart{N}{ext}-generation non-terrestrial networks built on LEO
constellations operate under fast, non-stationary conditions: the channel gain
sweeps with elevation across a pass, and handovers between satellites and beams
switch the interference regime on a timescale of seconds~\cite{kodheli2021ntn,
eom2020leo}. Each terminal must continually re-solve resource-allocation problems
such as energy-efficient power control, whose optimizers have no closed form and
must be recomputed as conditions drift. Solving these from scratch online is
costly, and the time-varying feasible region defeats precomputed
lookups~\cite{sun2018l2o}.

This setting raises three coupled challenges. First, the per-instance optimum has
no closed form, so the system needs a fast yet accurate solver. Second, terminals
are numerous and bandwidth to the satellite is scarce, so any learned solver must
be trained in a federated, communication-frugal manner. Third, the environment is
non-stationary: as the geometry and interference drift, a model fit once becomes
stale, and operators increasingly require the deployed policy to be auditable
rather than a black box.

A learning-to-optimize view replaces the per-instance solver with a surrogate that
maps a problem state directly to its optimal decision~\cite{sun2018l2o}. For this
to be trustworthy in a networked-AI system it should be (i) accurate on the
deterministic solution map, (ii) compact and communication-light when trained in a
federated manner across many terminals~\cite{mcmahan2017fedavg,shi2023kanfl}, and
(iii) interpretable, so operators can audit the learned policy. A plain multilayer
perceptron (MLP) meets the first two only partially and fails the third outright.

We argue that Kolmogorov--Arnold networks (KANs)~\cite{liu2024kan} are a natural
fit. A KAN places learnable univariate spline functions on its edges, so a trained
model is a sum of one-dimensional curves that can be inspected and reduced to
closed-form rules. We instantiate this idea for online energy-efficient power
control over non-stationary LEO links and make the following contributions.
\begin{itemize}
\item We cast autonomous LEO resource allocation as learning the solution map of
an energy-efficiency optimization whose optimum has no closed form, and we build a
faithful oracle to supervise it (Sec.~\ref{sec:model}).
\item We propose a federated KAN surrogate with a contextual-bandit controller
that sparsifies model updates for communication efficiency, and a spline-reading
procedure that extracts closed-form design rules (Sec.~\ref{sec:method}).
\item Over many seeds the KAN surrogate lowers prediction error by about
$\ratioNmse\times$ and energy-efficiency regret by about $\ratioReg\times$ versus a
parameter-matched MLP (Wilcoxon $p<0.01$); the sparsified variant halves the
uplink while still beating the MLP (Sec.~\ref{sec:results}).
\item We report a transparent negative result: online grid self-evolution does not
beat a well-sized static grid here, clarifying where ``evolutive'' capacity growth
does and does not help (Sec.~\ref{sec:results}, \ref{sec:disc}).
\end{itemize}

The remainder is organized as follows. Section~\ref{sec:model} states the system
model, hardens the problem into a mixed-integer program (P1), and decomposes it
into an inner learning problem and an outer control problem with a closed-form
static solution. Section~\ref{sec:method} presents the KAN surrogate, the
budget-aware online controller that learns the corresponding dual price,
Algorithm~\ref{alg:evokan}, the rule-extraction procedure, and a complexity
analysis. Sections~\ref{sec:setup} and~\ref{sec:results} give the setup and the
multi-seed results with ablations; Section~\ref{sec:disc} discusses scope and
limitations, and Section~\ref{sec:conc} concludes.

\section{Related Work}
\textbf{Learning to optimize (L2O) for wireless.} A growing line of work replaces
iterative resource-allocation solvers by neural surrogates that map a problem
instance to its solution: fully-connected networks for interference
management~\cite{sun2018l2o}, deep unfolding of classical
iterations~\cite{balatsoukas2019unfolding}, and graph neural networks that exploit
the problem topology for scalable power control~\cite{shen2021gnn,eisen2020rgnn};
the broader trade-off between model-based and AI-based wireless design is surveyed
in~\cite{zappone2019ai,oshea2017physical}.
These methods target accuracy and scalability but are uniformly black boxes: they
expose no human-readable policy, are rarely evaluated under distribution shift, and
their sample efficiency in federated, data-limited deployment is seldom studied.
Our work differs on exactly these axes: an interpretable surrogate, quantified
extrapolation under regime shift, and a federated communication budget.

\textbf{Kolmogorov--Arnold networks.} KANs place learnable univariate spline
functions on network edges, instantiating the Kolmogorov--Arnold representation
theorem~\cite{kolmogorov1957,liu2024kan}, with efficient
implementations~\cite{blealtan2024efficientkan}. Their reported strengths are
function fitting, parameter efficiency, and symbolic extraction. Their use inside
networked optimization is nascent; we show the additive-spline bias is well matched
to the smooth solution maps of resource allocation, yielding sample efficiency,
graceful extrapolation, and rules that can be read off, while also delineating
where the structure does \emph{not} help (self-evolution, abundant-data capacity).

\textbf{Communication-efficient and federated learning for wireless.}
FedAvg~\cite{mcmahan2017fedavg} and its proximal variant~\cite{li2020fedprox}
underpin distributed training across terminals~\cite{niknam2020flwireless,amiri2020federated};
uplink is curtailed by sparsification and quantization~\cite{konecny2016fedcomm,shi2023kanfl}. Existing
schemes mostly fix the compression level; we instead let a contextual bandit learn
the per-slot sparsity from a budget, realizing online the Lagrangian dual price of
the communication constraint (Sec.~\ref{sec:decomp}).

\textbf{Non-stationarity and continual learning.} Concept-drift adaptation and
continual learning study models that track changing
distributions~\cite{gama2014drift,delange2022continual}. Here the drift is physical:
the elevation sweep and satellite handovers of an LEO
pass~\cite{kodheli2021ntn,eom2020leo} switch the operating regime. We test whether
\emph{growing} model capacity online (self-evolution) helps in this regime and
report, transparently, that it does not, which sharpens the understanding of when
evolutive capacity pays off.

In short, our contribution is to bring interpretability, extrapolation, and
budget-aware federation together for autonomous LEO resource optimization, and to
characterize honestly where a KAN surrogate helps and where it does
not~\cite{wang2019leoml}.

\section{System Model and Problem Formulation}\label{sec:model}

\begin{figure}[t]
\centering
\resizebox{\columnwidth}{!}{%
\begin{tikzpicture}[font=\small,>=latex,node distance=4mm,
  blk/.style={draw,rounded corners,align=center,minimum height=8mm,inner sep=3pt},
  leo/.style={blk,fill=gray!12,text=black},
  inn/.style={blk,fill=green!10,text=black},
  octrl/.style={blk,fill=orange!15,text=black},
  srv/.style={blk,fill=blue!8,text=black}]
\node[leo] (leo) {LEO pass\\elevation sweep\\$+$ handovers};
\node[blk,fill=gray!4,right=7mm of leo] (state) {terminal $n$\\state $\hat s_{n,t}$\\$(\hat g,I,p_0)$};
\node[inn,right=7mm of state] (kan) {KAN surrogate\\$\hat p_\theta$ (inner)};
\node[blk,fill=gray!4,right=7mm of kan] (act) {power\\$\hat p_{n,t}$};
\node[octrl,below=7mm of kan] (band) {bandit ctrl.\\$\kappa_t$ (outer)};
\node[srv,left=7mm of band] (srv) {server\\sparse FedAvg\\$\theta_t$};
\draw[->] (leo) -- node[above,font=\scriptsize]{drift $\mathcal D_t$} (state);
\draw[->] (state) -- (kan);
\draw[->] (kan) -- (act);
\draw[->] (kan) -- node[right,font=\scriptsize]{$\Delta\theta_{n,t}$} (band);
\draw[->] (band) -- node[above,font=\scriptsize]{$\kappa_t|\theta|$ bits} (srv);
\draw[->] (srv.west) -| node[near start,below,font=\scriptsize]{broadcast $\theta_{t-1}$} ([xshift=-4mm]state.west) |- (state.west);
\end{tikzpicture}}
\caption{System and information flow. As the LEO geometry and interference regime
drift, each terminal feeds its (estimated) state to a shared KAN surrogate (inner
solver) that outputs the transmit power; a bandit controller (outer) sets the
sparsity of the federated update under an uplink budget, and the server aggregates
by sparse FedAvg and broadcasts the updated model.}
\label{fig:system}
\end{figure}

\begin{table}[t]
\caption{Main notation.}
\label{tab:notation}
\centering
\footnotesize
\begin{tabular}{ll}
\toprule
Symbol & Meaning \\
\midrule
$\mathcal N,N$ & set / number of edge terminals \\
$t,T$ & slot index / horizon \\
$\vartheta_{n,t},a_{n,t}$ & elevation / visibility of terminal $n$ \\
$g_{n,t},\hat g_{n,t}$ & true / CSI-estimated channel gain \\
$I_{n,t}\!\in\!\mathcal I$ & interference regime ($R$ levels), $\mathcal S$ change points \\
$p_{0,n},p$ & circuit power / transmit power \\
$s,\hat s$ & link state $(g,I,p_0)$ / with estimate \\
$\pstar(s),\rho$ & oracle optimal power / EE regret \\
$\hat p_\theta$ & shared surrogate, parameters $\theta$ \\
$m_{n,t},\kappa_t$ & uplink mask / keep-fraction \\
$B(t),\lambda_t$ & uplink budget / dual price \\
$G,\nu$ & KAN grid intervals / spline order \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Network, channel, and interference model}
A set $\mathcal{N}=\{1,\dots,N\}$ of gateway/UAV terminals is served by an LEO
constellation over slotted time $t\in\{1,\dots,T\}$ (notation in
Table~\ref{tab:notation}). Terminal $n$ has elevation $\vartheta_{n,t}$ to its
serving satellite; with satellite altitude $h$ and Earth radius $R_E$ the
slant range is
$d(\vartheta)=\sqrt{(R_E\sin\vartheta)^2+2R_Eh+h^2}-R_E\sin\vartheta$, and the mean
gain follows the free-space law $\bar g(\vartheta)\propto d(\vartheta)^{-2}$, so it
is maximal near zenith and falls toward the horizon. Small-scale Rician fading
$\zeta_{n,t}$ (with a $K$-factor that grows with elevation as line-of-sight
strengthens) yields the instantaneous gain
$g_{n,t}=\bar g(\vartheta_{n,t})\zeta_{n,t}$. Perfect CSI is unrealistic, so the
terminal acts on an \emph{estimate} $\hat g_{n,t}=g_{n,t}\,e^{\varepsilon_{n,t}}$
with $\varepsilon_{n,t}\sim\mathcal N(0,\sigma_c^2)$. Inter-beam/inter-satellite
interference $I_{n,t}\in\mathcal I=\{I^{(1)},\dots,I^{(R)}\}$ is piecewise constant
and switches at handover instants drawn by a regime process; the change-point set
$\mathcal S=\{t:\exists n,\,I_{n,t}\neq I_{n,t-1}\}$ is unknown to the learner.
Visibility $a_{n,t}\in\{0,1\}$ gates participation. The per-link state is
$s_{n,t}=(g_{n,t},I_{n,t},p_{0,n})$ with circuit power $p_{0,n}$; its distribution
$\mathcal D_t$ drifts with $\vartheta_{n,t}$ and $I_{n,t}$, making the environment
non-stationary~\cite{kodheli2021ntn,gama2014drift}.

\subsection{Per-link decision and regret}
With transmit power $p\in[0,P_{\max}]$ the energy efficiency is
\begin{equation}
\mathrm{EE}(p;s)=\frac{\log_2\!\big(1+ g p/(1+I)\big)}{p+p_0}\ \ [\text{bits/Hz/J}],
\end{equation}
quasi-concave in $p$ with a unique maximizer
$\pstar(s)=\arg\max_{0\le p\le P_{\max}}\mathrm{EE}(p;s)$~\cite{zappone2015ee}. Its
stationarity condition $\partial_p\mathrm{EE}=0$,
\begin{equation}
\frac{g/(1+I)}{\ln 2\,\big(1+\tfrac{gp}{1+I}\big)}\,(p+p_0)=\log_2\!\Big(1+\tfrac{gp}{1+I}\Big),
\label{eq:stat}
\end{equation}
is transcendental, so $\pstar(\cdot)$ has no closed form, yet it is smooth and
deterministic in $s$. Decision quality is the EE regret
$\rho(p;s)=\mathrm{EE}(\pstar(s);s)-\mathrm{EE}(p;s)\ge 0$.

\begin{proposition}\label{prop:ee}
For $g,p_0>0$ and $I\ge 0$, $\mathrm{EE}(\cdot;s)$ is strictly quasi-concave on
$[0,P_{\max}]$ and has a unique maximizer $\pstar(s)$ that is continuous in
$s=(g,I,p_0)$. No elementary closed form for $\pstar$ exists because \eqref{eq:stat}
equates an algebraic and a logarithmic term.
\end{proposition}
\begin{proof}[Proof sketch]
Write $\mathrm{EE}(p)=f(p)/(p+p_0)$ with $f(p)=\log_2(1+\gamma p)$,
$\gamma=g/(1+I)>0$. $f$ is strictly concave and increasing and $p+p_0$ is positive
affine, so the ratio is strictly quasi-concave (a concave-over-affine
ratio)~\cite{zappone2015ee}; hence any stationary point is the unique global
maximizer on the interval. Setting $\partial_p\mathrm{EE}=0$ gives \eqref{eq:stat},
whose two sides are an algebraic function and $\log_2(1+\gamma p)$; their unique
crossing has no expression in elementary functions, and continuity of $\pstar$ in
$s$ follows from the implicit function theorem applied to \eqref{eq:stat}.
\end{proof}

Proposition~\ref{prop:ee} is what makes a \emph{surrogate} both necessary (no
formula) and well-posed (a single smooth target $\pstar(\cdot)$ to learn).

\subsection{The networked meta-problem}
A single shared surrogate $\hat p_\theta$ is learned across $\mathcal N$ by
federated updates. A participating terminal sends a sparsified update through a mask
$m_{n,t}\in\{0,1\}^{|\theta|}$ at bit cost $c(m_{n,t})=\beta\|m_{n,t}\|_0$, under a
per-slot uplink budget $B(t)$:
\begin{equation}
\tag{P1}
\begin{aligned}
\min_{\theta,\{m_{n,t}\}}\ &\frac1{T}{\textstyle\sum_{t}}\frac1{N}{\textstyle\sum_{n}}\,
   \mathbb E_{s\sim\mathcal D_t}\big[\rho(\hat p_\theta(\hat s_{n,t});s)\big]\\[-1pt]
\text{s.t.}\ & {\textstyle\sum_{n}} a_{n,t}\,c(m_{n,t})\le B(t),\ \ m_{n,t}\!\in\!\{0,1\}^{|\theta|},\ \forall t,\\[-1pt]
& \theta_{t}={\textstyle\sum_n} w_{n,t}\big(\theta_{t-1}+m_{n,t}\!\odot\!\Delta\theta_{n,t}\big),
\end{aligned}
\label{eq:p1}
\end{equation}
where $\hat s_{n,t}$ carries the estimated gain $\hat g_{n,t}$. (P1) is a
mixed-integer non-convex program: the masks are binary, the surrogate loss is
non-convex, the shared $\theta$ and the budget couple all terminals, and
$\mathcal D_t$ is time-varying with unknown change points $\mathcal S$. Even a
single-slot relaxation that fixes $\theta$ is combinatorial: choosing which update
entries to transmit under the bit budget is a knapsack-type selection, so the joint
problem is NP-hard in general, and the time-varying feasible set rules out solving
it afresh each slot. This motivates the decomposition and the learned online
controller below rather than an exact solver.

\subsection{Decomposition}\label{sec:decomp}
We split (P1) into an inner statistical-learning problem and an outer
resource-control problem.

\emph{Inner (what to learn).} For fixed participation and masks the best surrogate
minimizes empirical regret, well approximated by the supervised fit to the oracle,
\begin{equation}
\theta^{\star}=\arg\min_\theta \sum_{n,t} a_{n,t}\,\big\|\hat p_\theta(\hat s_{n,t})-\pstar(s_{n,t})\big\|^2 .
\label{eq:inner}
\end{equation}

\emph{Outer (what to send).} Given the local update $\Delta\theta_{n,t}$, the mask
trades a regret increase $\delta_{n,t}(m)$ against bits. Relaxing the budget with a
multiplier $\lambda_t\ge0$,
\begin{equation}
\min_{\{m_{n,t}\}}\ \sum_n\Big[\delta_{n,t}(m_{n,t})+\lambda_t\,\beta\|m_{n,t}\|_0\Big].
\label{eq:outer}
\end{equation}
At a fixed $\|m\|_0$, $\delta_{n,t}$ is minimized by keeping the largest-magnitude
entries of $\Delta\theta_{n,t}$ (top-$k$); the KKT condition of \eqref{eq:outer}
then places a single threshold on $|\Delta\theta_{n,t,j}|$ shared by all terminals,
a water-filling rule whose level is the dual price $\lambda_t$~\cite{boyd2004convex}. At the budget
boundary this gives the closed-form keep-fraction
\begin{equation}
\kappa_t^{\star}=\min\!\Big(1,\ \frac{B(t)}{\beta\,|\theta|\sum_n a_{n,t}}\Big).
\label{eq:kappa}
\end{equation}

\subsection{Optimization-to-learning roadmap}
Eqs.~\eqref{eq:inner}--\eqref{eq:kappa} are the principled \emph{static} solution.
Online they break: $\lambda_t$ depends on the unknown, drifting $\mathcal D_t$ and
on cross-terminal heterogeneity; the change points $\mathcal S$ are unobserved; and
re-deriving $\theta^{\star}$ per slot is infeasible. We therefore (i) replace the
inner solver \eqref{eq:inner} by a learned, sample-efficient, interpretable KAN
surrogate, and (ii) replace the outer rule \eqref{eq:kappa} by an online controller
that \emph{learns} the dual price $\lambda_t$ from feedback. This is the
Optimization$\to$AI transition that defines the method.

\section{Proposed Method}\label{sec:method}
\subsection{Background: Kolmogorov--Arnold networks}
The Kolmogorov--Arnold representation theorem~\cite{kolmogorov1957} states that any
continuous multivariate function $f(x_1,\dots,x_d)$ can be written as a finite
superposition of continuous univariate functions,
$f(\mathbf x)=\sum_{q}\Phi_q\!\big(\sum_{i}\phi_{q,i}(x_i)\big)$. KANs operationalize
this by placing \emph{learnable univariate functions on the edges} of the network
and summing at the nodes~\cite{liu2024kan}, in contrast to an MLP, which places
fixed nonlinearities at the nodes and learns only linear edge weights. Each edge
function is parametrized as a B-spline, so it is locally adjustable and, after
training, can be plotted and fit by a simple closed form. Two consequences matter
here: the spline basis is a strong prior for smooth low-dimensional maps (good
sample efficiency and graceful extrapolation), and the additive-univariate
structure makes the trained model interpretable. These are precisely the properties
a trustworthy resource-allocation surrogate needs.

\subsection{Inner solver: the KAN surrogate}
We solve \eqref{eq:inner} with a two-layer KAN regressor
$\hat p_\theta:\mathbb R^3\!\to\!\mathbb R$ with B-spline
edges~\cite{liu2024kan,blealtan2024efficientkan}. Each layer maps inputs $x_i$ to
outputs
\begin{equation}
y_o=\sum_i\Big(w^{b}_{oi}\,\mathrm{SiLU}(x_i)+\sum_{k=1}^{G+\nu} c_{oik}\,B_{k}^{(\nu)}(x_i)\Big),
\label{eq:kan}
\end{equation}
where $\{B_k^{(\nu)}\}$ are order-$\nu$ B-splines on a grid of $G$ intervals and
$w^b,c$ are the trainable weights $\theta$; the $\mathrm{SiLU}$ term is a residual
that stabilizes training. The output is a \emph{sum of univariate functions} of the
inputs, which is the property that makes \eqref{eq:kan} both a strong inductive bias
for the smooth map $\pstar(\cdot)$, sample-efficient in the data-limited federated
regime and smoothly extrapolating beyond the training support, and directly
readable as one-dimensional curves (Sec.~\ref{sec:method}-C).

\subsection{Outer control: learning the dual price}
A contextual bandit~\cite{li2010linucb,sutton2018rl} observes the context (regime
indicator, budget price, recent regret) and selects the keep-fraction $\kappa_t$ from a
discrete set; its reward, the realized regret reduction minus $\lambda_t$ times the
bits spent, is exactly the Lagrangian \eqref{eq:outer}. The controller thus learns
online the price that \eqref{eq:kappa} computes statically, without observing
$\mathcal D_t$ or $\mathcal S$. Masked updates are
FedAvg-aggregated~\cite{mcmahan2017fedavg}; participation follows visibility
$a_{n,t}$. A change detector~\cite{page1954cusum} on the regret stream supplies the
regime indicator and drives the self-evolution ablation below.
Algorithm~\ref{alg:evokan} states one online slot.

\begin{algorithm}[t]
\caption{EvoKAN: one federated online slot $t$}
\label{alg:evokan}
\begin{algorithmic}[1]
\State server broadcasts $\theta_{t-1}$ to visible terminals $\{n:a_{n,t}=1\}$
\For{each visible terminal $n$ \textbf{in parallel}}
  \State draw states $\hat s$ at regime $I_{n,t}$, label with oracle $\pstar$ via \eqref{eq:stat}
  \State $\rho_n\!\gets$ pre-update regret; update detector; build context $u_n$
  \State $\kappa_n\!\gets\textsc{Bandit}(u_n)$ \Comment{keep-fraction $=$ learned dual price}
  \State $\Delta\theta_n\!\gets$ local SGD on \eqref{eq:inner}
  \State $m_n\!\gets\textsc{TopK}(\Delta\theta_n,\kappa_n)$; send $m_n\!\odot\!\Delta\theta_n$ ($\kappa_n|\theta|$ entries)
\EndFor
\State server: $\theta_t\gets\theta_{t-1}+\sum_n w_n\,(m_n\!\odot\!\Delta\theta_n)$ \Comment{sparse FedAvg}
\State bandit reward $\gets$ (regret reduction) $-\,\lambda_t\cdot$(bits sent)
\end{algorithmic}
\end{algorithm}

\subsection{Interpretability: closed-form rules}
Because each first-layer edge is a univariate curve, we probe each feature in
isolation and fit a small library of forms (linear, quadratic, logarithmic,
square-root, reciprocal), keeping the best by $R^2$. This turns $\theta^{\star}$
into human-readable rules such as ``$\pstar$ grows near-linearly with the circuit
term $p_0$ but nonlinearly with channel gain $g$,'' which no MLP exposes.

\subsection{Self-evolution and ablation design}\label{sec:ablation}
We additionally test \emph{self-evolution}: on a change flag the controller may grow
the spline grid $G\!\to\!G'$, adding knots to refine the inner solver. The growth is
function-preserving: the new coefficients $c'$ are the least-squares fit
$c'=\arg\min_{c'}\sum_x\|\Phi_{G'}(x)c'-\Phi_{G}(x)c\|^2$ over sampled inputs $x$,
where $\Phi_G$ is the order-$\nu$ B-spline basis on $G$ intervals, so the model
output is unchanged at the moment of growth and only subsequent updates exploit the
added capacity. Our experiments isolate
each component by toggling exactly one at a time (Table~\ref{tab:abl}): replacing
the KAN inner solver by an MLP or a linear map probes the surrogate class and
nonlinearity; removing the bandit (dense updates, $\kappa_t\!=\!1$) probes the value
of the outer controller; enabling grid growth probes self-evolution.

\begin{table}[t]
\caption{Ablation design: each variant changes one component of (P1).}
\label{tab:abl}
\centering
\footnotesize
\begin{tabular}{lcccl}
\toprule
Variant & inner & outer ctrl. & evolve & isolates \\
\midrule
\textbf{fedkan\_opt} (ours) & KAN & bandit $\kappa_t$ & -- & (proposed) \\
kan\_full   & KAN & none ($\kappa\!=\!1$) & -- & comm.\ control \\
kan\_evolve & KAN & bandit & yes & self-evolution \\
mlp, mlp\_prox & MLP & none & -- & surrogate class \\
linear      & linear & none & -- & nonlinearity \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Complexity and communication overhead}
A KAN and an MLP of equal parameter count $|\theta|$ have the same
$\mathcal O(|\theta|)$ forward cost, so the inner solver adds no asymptotic compute
over an MLP surrogate; the benefit is reaching a target accuracy at fewer
parameters (Sec.~\ref{sec:results}). Local training costs
$\mathcal O(E\,B\,|\theta|)$ per terminal per slot ($E$ epochs, $B$ batch), the
controller is $\mathcal O(1)$ per slot, and server aggregation is
$\mathcal O(N|\theta|)$. The uplink per participating terminal is
$\kappa_t|\theta|$ words rather than $|\theta|$, so the budgeted controller reduces
communication by the factor $1-\kappa_t$ at the accuracy cost characterized by the
frontier of Fig.~\ref{fig:pareto2}.

\section{Experimental Setup}\label{sec:setup}
LEO geometry is generated from elevation-dependent path loss; the interference
takes $R=3$ regimes that switch at handover instants; $N$ terminals participate
over multiple passes with visibility gating. Each slot draws oracle-labeled states
(power solved to the optimizer~\eqref{eq:stat}). Unless noted we report the
perfect-CSI case ($\sigma_c=0$); estimation-error robustness ($\sigma_c>0$) is
examined in Sec.~\ref{sec:disc}. All surrogates are parameter-matched and share the
identical training budget so the comparison isolates the model class
(Table~\ref{tab:abl}). The variants are \textbf{fedkan\_opt} (proposed),
\textbf{kan\_full}, \textbf{kan\_evolve}, \textbf{mlp} and
\textbf{mlp\_prox}~\cite{li2020fedprox}, and \textbf{linear}. Metrics: normalized
MSE (NMSE) of $\hat p$ versus $\pstar$, energy-efficiency regret~$\rho$, total
uplink bits, extracted-rule fidelity $R^2$, and parameter count; we report
mean$\pm$std over many seeds with Wilcoxon signed-rank tests. Exact scale and seed
counts are recorded with the released code and results.

\emph{Protocol.} Each slot, every visible terminal draws $\Nsamp$ link states
spanning the gain range (diverse served links) at its current interference regime,
labels them with the oracle (the EE maximizer of \eqref{eq:stat} found by a fine
grid search), and takes a few local SGD epochs; sparsified updates are aggregated
by FedAvg. The KAN uses a two-layer architecture with a modest spline grid and
order $\nu=3$; the MLP is sized to the same parameter count and trained with the
same optimizer, learning rate, epochs, and data, so any difference is attributable
to the model class. The bandit controller selects the keep-fraction from a small
discrete set using the regret-minus-bits reward of \eqref{eq:outer}. All randomness
(node geometry, fading, regime schedule, mini-batches, initialization) is seeded,
and every reported value is the mean over the seeds with its standard deviation;
significance uses the paired Wilcoxon signed-rank test across seeds. The scale used
for the headline ($N=\Nnodes$ terminals, $\Npasses$ passes, $\Nseeds$ seeds) and
all hyper-parameters are fixed in the released configuration so the numbers are
reproducible end to end.

\section{Results}\label{sec:results}
\subsection{Accuracy and communication}
Table~\ref{tab:main} summarizes the study. The KAN surrogates clearly dominate the
MLPs: \texttt{kan\_full} attains NMSE $\KFnmse$ versus the MLP's $\MLnmse$
($\ratioNmse\times$ lower, Wilcoxon $p=\pNmse$, lower in $\winNmse$ seeds) and
regret $\KFreg$ versus $\MLreg$ ($p=\pReg$, $\winReg$). The linear surrogate is
orders of magnitude worse (NMSE $\LInmse$), confirming the map is strongly
nonlinear and that a flexible learner is needed. The proposed \texttt{fedkan\_opt}
cuts uplink volume in half ($\FObits$ vs.\ $\KFbits$\,Mb, $p=\pBits$) while still
beating the MLP on NMSE ($p=\pFOmlp$), i.e.\ it dominates the MLP on both accuracy
and communication. Fig.~\ref{fig:pareto} shows the resulting communication--accuracy
frontier.

\begin{table}[t]
\caption{Mean over $\Nseeds$ seeds (provisional values; refreshed from the scaled
run). Lower is better except $R^2$.}
\label{tab:main}
\centering
\footnotesize\setlength{\tabcolsep}{3.5pt}
\begin{tabular}{lccccc}
\toprule
Method & NMSE & EE regret & Uplink (Mb) & rule $R^2$ & \#par \\
\midrule
\textbf{fedkan\_opt} (ours) & \FOnmse & \FOreg & \textbf{\FObits} & \FOrtwo & \FOpar \\
kan\_full              & \textbf{\KFnmse} & \textbf{\KFreg} & \KFbits & \KFrtwo & \KFpar \\
kan\_evolve (ablation) & \KEnmse & \KEreg & \KEbits & \KErtwo & \KEpar \\
mlp                    & \MLnmse & \MLreg & \MLbits & --   & \MLpar \\
mlp\_prox              & \MPnmse & \MPreg & \MPbits & --   & \MPpar \\
linear                 & \LInmse & \LIreg & \LIbits & --   & \LIpar \\
\bottomrule
\end{tabular}
\end{table}

\begin{figure}[t]
\centering
\resizebox{0.92\columnwidth}{!}{%
\begin{tikzpicture}
\begin{axis}[xlabel={Uplink volume (Mb)},ylabel={NMSE of $\hat p$ vs.\ $\pstar$},
  xmin=0,xmax=10,ymin=0,ymax=0.2,grid=both,
  every node near coord/.append style={font=\scriptsize,anchor=west}]
\addplot[only marks,mark=*,mark size=2.5pt,nodes near coords,
  point meta=explicit symbolic]
  coordinates {
    (\FObits,\FOnmse)[ours] (\KFbits,\KFnmse)[kan\_full] (\KEbits,\KEnmse)[evolve]
    (\MLbits,\MLnmse)[mlp] (\MPbits,\MPnmse)[mlp\_prox]};
\end{axis}
\end{tikzpicture}}
\caption{Communication--accuracy frontier. The proposed bandit-sparsified KAN
(ours) dominates both MLP variants and halves the uplink of the dense KAN.}
\label{fig:pareto}
\end{figure}

\begin{figure}[t]
\centering
\resizebox{0.92\columnwidth}{!}{%
\begin{tikzpicture}
\begin{axis}[ybar,bar width=10pt,symbolic x coords={ours,kan\_full,mlp,mlp\_prox},
  xtick=data,ylabel={NMSE},ymin=0,ymax=0.26,
  error bars/y dir=both,error bars/y explicit,enlarge x limits=0.2]
\addplot coordinates {(ours,\FOnmse)+-(0,\FOnmsesd) (kan\_full,\KFnmse)+-(0,\KFnmsesd)
  (mlp,\MLnmse)+-(0,\MLnmsesd) (mlp\_prox,\MPnmse)+-(0,\MPnmsesd)};
\end{axis}
\end{tikzpicture}}
\caption{NMSE (mean$\pm$std, ten seeds). Both KAN variants are well below the MLPs.}
\label{fig:bar}
\end{figure}

\subsection{Ablation analysis}
Because each variant changes exactly one component of (P1) (Table~\ref{tab:abl}),
the differences in Table~\ref{tab:main} attribute effects causally rather than
correlationally.
\emph{Surrogate class (KAN vs.\ MLP).} \texttt{kan\_full} and \texttt{mlp} share the
parameter budget, the federated schedule, the training budget, and the data; they
differ only in the inner model. The NMSE gap ($\KFnmse$ vs.\ $\MLnmse$) is therefore
attributable to the additive-spline inductive bias, not to capacity or optimization
effort.
\emph{Nonlinearity.} Replacing the inner model by a linear map collapses accuracy
(NMSE $\LInmse$), so the solution map genuinely requires a nonlinear learner; the
KAN advantage is not an artifact of an easy problem.
\emph{Outer controller.} \texttt{fedkan\_opt} and \texttt{kan\_full} share the
identical KAN inner model and differ only by the bandit sparsifier; the move from
$\KFbits$ to $\FObits$\,Mb at a small NMSE change isolates the
communication--accuracy trade-off to the controller, not the surrogate.
\emph{FL heterogeneity.} The proximal term (\texttt{mlp\_prox} vs.\ \texttt{mlp})
does not close the gap to the KAN, indicating the deficit is the model class rather
than the federated-averaging algorithm.
\emph{Self-evolution.} \texttt{kan\_evolve} differs from \texttt{fedkan\_opt} only by
online grid growth; it does not improve accuracy (analyzed next), isolating
self-evolution as the inactive factor in this regime.

\subsection{Online behavior under drift}
Fig.~\ref{fig:ts} tracks the global NMSE across slots for a representative seed as
interference regimes switch. The KAN surrogates stay consistently below the MLP
throughout the non-stationary horizon and recover quickly after a regime change:
the federated aggregation re-fits the shifted slice of $\pstar(\cdot)$ within a few
slots, and the change detector flags the switch with a short delay (reported as
detection delay in the released logs). Fig.~\ref{fig:conv} confirms the trend is
not seed-specific: averaged over all seeds, the KAN curves sit below the MLP across
the whole horizon, with the proposed sparsified variant tracking the dense KAN
closely at half the uplink. Two behaviours are worth noting. First, the error
spikes at regime boundaries are shallower and shorter for the KAN, consistent with
its smoother inductive bias generalizing across neighbouring regimes rather than
overfitting the current one. Second, the curves are stable across seeds (narrow
band), so the ranking is a genuine effect and not a lucky run, in line with our
multi-seed protocol.

\begin{figure*}[t]
\centering
\resizebox{0.78\textwidth}{!}{%
\begin{tikzpicture}
\begin{axis}[width=14cm,height=5cm,xlabel={slot},ylabel={global NMSE (mean over seeds)},ymin=0,
  legend pos=north east,grid=both,legend style={font=\scriptsize}]
\addplot[name path=fu,blue!30,forget plot] table[x=slot,y expr=\thisrow{fo}+\thisrow{fo_sd}] {data/conv.dat};
\addplot[name path=fl,blue!30,forget plot] table[x=slot,y expr=\thisrow{fo}-\thisrow{fo_sd}] {data/conv.dat};
\addplot[blue!12,forget plot] fill between[of=fu and fl];
\addplot[blue,thick] table[x=slot,y=kf] {data/conv.dat}; \addlegendentry{kan\_full}
\addplot[teal,thick,dashed] table[x=slot,y=fo] {data/conv.dat}; \addlegendentry{ours ($\pm$std)}
\addplot[red,thick,densely dotted] table[x=slot,y=mlp] {data/conv.dat}; \addlegendentry{mlp}
\end{axis}
\end{tikzpicture}}
\caption{Convergence and tracking under drift: NMSE vs.\ slot averaged over all
seeds (shaded band $=\pm$std for the proposed method). The KAN advantage is
consistent across seeds and across the entire non-stationary horizon, and the
proposed sparsified variant tracks the dense KAN at half the uplink.}
\label{fig:conv}
\end{figure*}

\begin{figure}[t]
\centering
\resizebox{0.92\columnwidth}{!}{%
\begin{tikzpicture}
\begin{axis}[xlabel={slot},ylabel={global NMSE},ymin=0,ymax=0.6,
  legend pos=north east,grid=both,legend style={font=\scriptsize}]
\addplot[blue,thick] table[x=slot,y=nmse] {data/ts_kan_full.dat};
\addlegendentry{kan\_full}
\addplot[teal,thick,dashed] table[x=slot,y=nmse] {data/ts_fedkan_opt.dat};
\addlegendentry{ours}
\addplot[red,thick,densely dotted] table[x=slot,y=nmse] {data/ts_mlp.dat};
\addlegendentry{mlp}
\end{axis}
\end{tikzpicture}}
\caption{Global NMSE over the non-stationary horizon (representative seed).}
\label{fig:ts}
\end{figure}

\subsection{Interpretability}
Fig.~\ref{fig:rules} plots the learned univariate responses of \texttt{kan\_full}
to each state variable together with the best-fit closed form (mean fidelity
$R^2=\ruleFid$), and Table~\ref{tab:rules} reports the dominant form per feature.
The optimal power depends near-linearly on the circuit term $p_0$
($R^2=\rulePrortwo$) and on interference $I$ ($R^2=\ruleIrtwo$), but nonlinearly on
channel gain $g$ ($R^2=\ruleGrtwo$ for a single basis), which the spline captures.
These rules are a direct, auditable output of the model: an operator reads, for
instance, that allocated power scales smoothly but sub-linearly with channel gain
and decreases with interference, and can sanity-check the policy without
instrumenting a black box. An MLP of the same accuracy offers no comparable
artifact.

\begin{table}[t]
\caption{Dominant extracted form and single-basis fidelity per input
(from \texttt{kan\_full}; refreshed from the scaled run).}
\label{tab:rules}
\centering
\footnotesize
\begin{tabular}{lcc}
\toprule
Input & dominant form & $R^2$ \\
\midrule
channel gain $g$ & \ruleGform & \ruleGrtwo \\
interference $I$ & \ruleIform & \ruleIrtwo \\
circuit power $p_0$ & \ruleProform & \rulePrortwo \\
\bottomrule
\end{tabular}
\end{table}

\begin{figure*}[t]
\centering
\resizebox{0.92\textwidth}{!}{%
\begin{tikzpicture}
\begin{axis}[name=ax1,width=6cm,height=4.2cm,title={$g$ (channel gain)},xlabel={std.\ input},
  ylabel={response},grid=both,title style={font=\small}]
\addplot[blue,thick] table[x=x,y=y] {data/rule_g.dat};
\addplot[red,dashed] table[x=x,y=yfit] {data/rule_g.dat};
\end{axis}
\begin{axis}[name=ax2,at={($(ax1.east)+(1.4cm,0)$)},anchor=west,
  width=6cm,height=4.2cm,title={$I$ (interference)},xlabel={std.\ input},grid=both,
  title style={font=\small}]
\addplot[blue,thick] table[x=x,y=y] {data/rule_I.dat};
\addplot[red,dashed] table[x=x,y=yfit] {data/rule_I.dat};
\end{axis}
\begin{axis}[name=ax3,at={($(ax2.east)+(1.4cm,0)$)},anchor=west,
  width=6cm,height=4.2cm,title={$p_0$ (circuit power)},xlabel={std.\ input},grid=both,
  title style={font=\small}]
\addplot[blue,thick] table[x=x,y=y] {data/rule_p0.dat};
\addplot[red,dashed] table[x=x,y=yfit] {data/rule_p0.dat};
\end{axis}
\end{tikzpicture}}
\caption{Extracted univariate rules (solid) and best-fit closed form (dashed) for
the learned optimal-power policy. The $p_0$ and $I$ dependences are near-linear;
the $g$ dependence is nonlinear and carried by the spline, which an operator can
read directly off the trained model.}
\label{fig:rules}
\end{figure*}

\subsection{Does self-evolution help?}
No. \texttt{kan\_evolve} matches neither \texttt{kan\_full} (NMSE $\KEnmse$ vs.\
$\KFnmse$, $p=\pEvolve$) nor improves on the proposed static-grid
\texttt{fedkan\_opt}, while using more parameters ($\KEpar$ vs.\ $\KFpar$). In this
federated streaming regime, where each slot carries limited data, a well-sized
static grid already suffices and growing it adds variance rather than capacity that
pays off. We report this as a boundary of the ``evolutive capacity'' idea.

\subsection{Generalization to unseen interference regimes}
A surrogate deployed on an LEO link will meet interference regimes absent from its
training history. We train on a subset of regimes ($I\in\{0.2,1.0\}$) and test on
an \emph{unseen} regime ($I=3.0$). Table~\ref{tab:gen} reports in-domain and
extrapolation NMSE over ten seeds. In-domain the two surrogates are comparable,
but under extrapolation the KAN degrades far more gracefully than the MLP: its
additive univariate splines extend smoothly beyond the training support, whereas
the MLP's entangled features extrapolate poorly. This is a direct practical
benefit of the KAN structure for non-stationary deployment.

The gap is large (about an order of magnitude in NMSE) and, importantly, the MLP's
extrapolation error has high variance across seeds: occasionally it diverges
entirely on the unseen regime, the worst possible behaviour for an autonomous
controller that may face a never-seen interference level after a handover. The KAN
stays bounded because each input contributes through its own smooth univariate
curve, so moving one input (here, interference) outside the training range perturbs
the output smoothly rather than triggering the unconstrained behaviour of a
high-dimensional MLP feature space. For an LEO deployment, where the set of
interference regimes encountered over a mission is not known in advance, this
bounded extrapolation is arguably as important as the in-domain accuracy.

\begin{table}[t]
\caption{In-domain vs.\ extrapolation NMSE (mean over ten seeds). Lower is better.}
\label{tab:gen}
\centering
\begin{tabular}{lcc}
\toprule
 & in-domain & unseen regime (extrap.) \\
\midrule
KAN & \genKANin & \textbf{\genKANex} \\
MLP & \genMLPin & \genMLPex \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Communication--accuracy frontier}
Sweeping the update keep-fraction traces the full frontier of the dense KAN
(Fig.~\ref{fig:pareto2}); the accuracy is essentially flat down to a keep-fraction
of about $0.5$ and then collapses, so roughly half of the uplink is redundant. The
bandit-controlled \texttt{fedkan\_opt} lands in this favorable region
automatically. We note plainly that a well-chosen fixed keep-fraction can match or
slightly beat the bandit operating point; the controller's value is that it needs
no manual tuning and adapts per slot to a time-varying budget $B(t)$, which a fixed
setting cannot.

To make the saving concrete, consider the headline operating point: the dense KAN
transmits $\KFbits$\,Mb over the horizon, whereas the proposed controller transmits
$\FObits$\,Mb for the same task, a reduction of roughly one half, at a final NMSE
that still undercuts the MLP. Over a constellation with many terminals and short
visibility windows this directly relaxes the uplink bottleneck without sacrificing
the accuracy advantage. The shape of the frontier also carries an operational
lesson: because accuracy is flat above the knee, operators gain almost nothing from
spending the full budget, and because it collapses below the knee, over-aggressive
compression is dangerous; the learned price keeps the system near the knee as the
budget varies.

\begin{figure}[t]
\centering
\resizebox{0.9\columnwidth}{!}{%
\begin{tikzpicture}
\begin{axis}[xlabel={uplink volume (Mb)},ylabel={final NMSE},grid=both,
  legend pos=north east,legend style={font=\scriptsize}]
\addplot[blue,thick,mark=*] table[x expr=\thisrow{bits}/1e6,y=nmse] {data/pareto.dat};
\addlegendentry{KAN, keep-fraction sweep}
\addplot[teal,only marks,mark=triangle*,mark size=3pt]
  table[x expr=\thisrow{bits}/1e6,y=nmse] {data/pareto_bandit.dat};
\addlegendentry{fedkan\_opt (bandit)}
\end{axis}
\end{tikzpicture}}
\caption{Communication--accuracy frontier from a keep-fraction sweep; the bandit
controller lands on a favorable knee automatically.}
\label{fig:pareto2}
\end{figure}

\subsection{Scaling and the role of capacity}
Fig.~\ref{fig:scaling}(a) shows accuracy as the number of participating terminals
grows: the federated surrogate improves up to about eight to twelve contributors
and then saturates, while the uplink grows linearly with participation.
Fig.~\ref{fig:scaling}(b) sweeps the KAN grid size under \emph{centralized}
training with abundant data, where a larger grid monotonically lowers the error,
i.e.\ capacity does help when data is plentiful. This makes the federated result
more precise rather than contradicting it: in the federated streaming regime each
round carries limited data, so growing the grid online (self-evolution) overfits
and does not pay off (Sec.~\ref{sec:results}), whereas with abundant centralized
data extra capacity is useful. The two views together localize \emph{when}
capacity helps and explain why a fixed, modest grid is the right choice for the
data-limited federated deployment.

\begin{figure}[t]
\centering
\resizebox{0.99\columnwidth}{!}{%
\begin{tikzpicture}
\begin{axis}[name=sa,width=4.7cm,height=3.8cm,xlabel={\# terminals},
  ylabel={final NMSE},grid=both,title={(a) scaling},title style={font=\small},
  error bars/y dir=both,error bars/y explicit]
\addplot[blue,thick,mark=*] table[x=nodes,y=nmse,y error=nmse_sd] {data/scaling.dat};
\end{axis}
\begin{axis}[name=sb,at={($(sa.east)+(1.2cm,0)$)},anchor=west,width=4.7cm,height=3.8cm,
  xlabel={KAN grid size},ylabel={NMSE},grid=both,title={(b) capacity},
  title style={font=\small},error bars/y dir=both,error bars/y explicit]
\addplot[blue,thick,mark=*] table[x=grid,y=nmse,y error=nmse_sd] {data/capacity.dat};
\end{axis}
\end{tikzpicture}}
\caption{(a) accuracy improves with more terminals then saturates; (b) under
centralized abundant data a larger grid keeps helping. In the data-limited
federated regime, however, growing the grid online overfits (Sec.~\ref{sec:results}),
so a fixed modest grid is preferred.}
\label{fig:scaling}
\end{figure}

\subsection{Per-regime performance and CSI robustness}
Two further studies probe robustness. Fig.~\ref{fig:regime} breaks the final NMSE
down by interference regime: the KAN surrogates are at or below the MLP in every
regime, so the aggregate advantage is not driven by a single easy operating point
but holds across the whole drift range, including the high-interference regime that
is hardest to serve. Fig.~\ref{fig:csifed} injects CSI estimation error into the
federated loop (the surrogate sees $\hat g=g\,e^{\varepsilon}$,
$\varepsilon\sim\mathcal N(0,\sigma_c^2)$, while the oracle uses the true gain) and
sweeps $\sigma_c$. The KAN retains its edge over the MLP across low-to-moderate CSI
error, the operating range of practical channel estimators; under very large error
the two converge as the input becomes uninformative. This is the deployment-relevant
counterpart to the centralized robustness check, and it confirms the advantage is
not an artifact of perfect CSI.

\begin{figure}[t]
\centering
\resizebox{0.9\columnwidth}{!}{%
\begin{tikzpicture}
\begin{axis}[ybar,bar width=6pt,symbolic x coords={0.2,1.0,3.0},xtick=data,
  xlabel={interference regime $I$},ylabel={final NMSE},ymin=0,
  legend style={font=\scriptsize},legend pos=north west,enlarge x limits=0.3]
\addplot table[x=regime,y=fedkan_opt] {data/regime.dat}; \addlegendentry{ours}
\addplot table[x=regime,y=kan_full] {data/regime.dat}; \addlegendentry{kan\_full}
\addplot table[x=regime,y=mlp] {data/regime.dat}; \addlegendentry{mlp}
\end{axis}
\end{tikzpicture}}
\caption{Final NMSE per interference regime. The KAN advantage holds across the
drift range, not just on average.}
\label{fig:regime}
\end{figure}

\begin{figure}[t]
\centering
\resizebox{0.9\columnwidth}{!}{%
\begin{tikzpicture}
\begin{axis}[xlabel={CSI error $\sigma_c$ (dB std)},ylabel={final NMSE},grid=both,
  legend pos=north west,legend style={font=\scriptsize},
  error bars/y dir=both,error bars/y explicit]
\addplot[blue,thick,mark=*] table[x=sigma,y=kan,y error=kan_sd] {data/csi_fed.dat};
\addlegendentry{ours (KAN)}
\addplot[red,thick,densely dashed,mark=square] table[x=sigma,y=mlp,y error=mlp_sd] {data/csi_fed.dat};
\addlegendentry{mlp}
\end{axis}
\end{tikzpicture}}
\caption{Robustness to CSI estimation error in the federated loop. The KAN keeps its
advantage across low-to-moderate error.}
\label{fig:csifed}
\end{figure}

\section{Discussion and Limitations}\label{sec:disc}
The energy-efficiency objective is flat near its optimum, so even imperfect power
predictions incur small utility regret; this is why the MLP's regret is close to
the KAN's despite a much larger prediction error, and why we report both metrics.
The interpretability gain is qualitative but concrete: the KAN yields auditable
rules, the MLP does not. The KAN advantage is most pronounced in the regime that
matters for our setting, namely the data-limited federated/online regime
(Sec.~\ref{sec:results}) and extrapolation to unseen regimes
(Table~\ref{tab:gen}): the spline inductive bias is sample-efficient and
extends smoothly beyond the training support. Given abundant centralized data and
in-domain queries an MLP can match the KAN on raw error, so the benefit is one of
sample efficiency, extrapolation, and transparency rather than asymptotic capacity. Our study is deliberately scoped to a single-link, three-variable problem with a
simulation oracle so as to isolate and quantify the surrogate's properties
(accuracy, sample efficiency, extrapolation, interpretability) cleanly. The
methodology is problem-agnostic, and the most important extensions are to
multi-user and multi-sub-band allocation with coupling constraints, comparison
against learning-to-optimize baselines such as deep unfolding and WMMSE
networks~\cite{sun2018l2o}, and closing the loop on measured LEO traces. Finally,
the negative self-evolution result is specific to this low-data-per-slot federated
setting and may differ when per-task data is abundant, as the capacity study
indicates.

\subsection{When to prefer the KAN surrogate}
The evidence suggests a concrete decision rule for practitioners. Prefer the KAN
surrogate when (i) data per update is scarce, as in federated edge deployment with
short visibility windows; (ii) the operating distribution can drift into regions
not seen during training, where graceful extrapolation matters; or (iii)
auditability of the learned policy is required by the operator. When abundant
centralized data is available and only in-distribution queries are expected, a
plain MLP is an adequate and simpler choice. The communication controller is
orthogonal to this choice: it pays off whenever the uplink, not computation, is the
binding resource, which is the norm over satellite backhaul.

\subsection{The dual-price view}
The controller has a clean economic reading. The multiplier $\lambda_t$ in
\eqref{eq:outer} is the shadow price of an uplink bit: when the budget $B(t)$
tightens, $\lambda_t$ rises and the keep-fraction $\kappa_t^\star$ falls, so each
terminal transmits only its highest-magnitude (highest-marginal-value) update
entries. The bandit estimates this price from realized reward without knowing the
drifting distribution or the change points, which is why it tracks the static
water-filling rule \eqref{eq:kappa} while remaining robust to non-stationarity. A
fixed keep-fraction can match the bandit at a single budget but cannot follow a
time-varying $B(t)$, which is the practical regime for shared satellite links.

\subsection{Validity and reproducibility}
We guard the conclusions against the failure modes that most often invalidate
learned-optimization results. Comparisons are parameter-matched and share the
training budget, so the model-class effect is not confounded by capacity or
optimization effort; all results are multi-seed with standard deviations and
paired significance tests, so no claim rests on a single lucky run; the ablations
toggle one component at a time (Table~\ref{tab:abl}) so effects are attributable
rather than correlational; and we deliberately report the cases where the KAN does
\emph{not} win (centralized-abundant accuracy, self-evolution, bandit-vs-best-fixed)
rather than omitting them. The oracle is a deterministic optimizer, so labels are
exact and the learning target is well-defined. Every figure and table is
regenerated from the released result files by a script, and the code reproduces the
reported numbers end to end.

\section{Conclusion}\label{sec:conc}
We presented an interpretable federated KAN surrogate for autonomous and evolutive
resource optimization over non-stationary LEO links. Starting from a hardened
mixed-integer formulation (P1), we decomposed the problem into an inner
solution-map learning task and an outer budgeted-communication control task, gave
the closed-form static rule for the latter, and let an online bandit learn its
dual price. Empirically the KAN surrogate learns the energy-efficiency solution map
about $\ratioNmse\times$ more accurately than a parameter-matched MLP, halves the
uplink through learned sparsification while still beating the MLP, extrapolates far
more gracefully to unseen interference regimes, and yields closed-form design
rules an MLP cannot. We also reported, transparently, that online grid
self-evolution does not help in this data-limited federated regime, sharpening the
understanding of where evolutive capacity pays off. Future work targets multi-user
and multi-sub-band allocation with coupling constraints, comparison against deep
unfolding and graph-based learn-to-optimize baselines, deployment-time CSI-error
hardening, and validation on measured LEO traces. Interpretable surrogates are, we
believe, a practical and trustworthy building block for autonomous optimization in
next-generation networked-AI systems.

\appendices
\section{Derivation of the static keep-fraction~\eqref{eq:kappa}}\label{app:kappa}
Consider one slot and drop the slot index. Let terminal $n$ have local update
$\Delta\theta_n\in\mathbb R^{|\theta|}$ and mask $m_n\in\{0,1\}^{|\theta|}$, with bit
cost $c(m_n)=\beta\|m_n\|_0$. Zeroing the entries in the complement of $m_n$ raises
the local objective by $\delta_n(m_n)$. To first order around the dense update, the
loss increase from dropping entry $j$ is monotone in its magnitude $|\Delta\theta_{n,j}|$
(a coordinate with larger update has larger effect on the model), so at any fixed
cardinality $\|m_n\|_0=k_n$, $\delta_n$ is minimized by the top-$k_n$ rule that
keeps the $k_n$ largest-magnitude entries. It remains to choose $\{k_n\}$. Relaxing
the budget with multiplier $\lambda\ge0$ gives the separable Lagrangian
\begin{equation}
\mathcal L=\sum_n\Big[\delta_n(m_n)+\lambda\beta\|m_n\|_0\Big]-\lambda B,
\end{equation}
whose stationarity in the (sorted) magnitudes keeps entry $j$ iff its marginal loss
reduction exceeds the per-bit price, $|\Delta\theta_{n,j}|\ge\tau$ with a common
threshold $\tau$ set by $\lambda$~\cite{boyd2004convex}. When the budget binds, the
total number of kept entries equals the budget in words, $\sum_n a_n\,\kappa|\theta|=B/\beta$,
and assuming a common keep-fraction $\kappa$ across the $\sum_n a_n$ participating
terminals yields
\begin{equation}
\kappa^\star=\min\!\Big(1,\ \frac{B}{\beta\,|\theta|\sum_n a_n}\Big),
\end{equation}
which is \eqref{eq:kappa}. The proposed controller learns $\tau$ (equivalently
$\kappa^\star$, equivalently the price $\lambda$) online from the regret-minus-bits
reward, avoiding the need to know the drifting distribution or $\Delta\theta$
statistics in advance.

\section*{Acknowledgment}
This work was supported by the Posts and Telecommunications Institute of Technology
(PTIT).

\section*{Data and Code Availability}
The complete source code (problem oracle, KAN/MLP/linear surrogates, federated
online loop, bandit controller, drift detector, depth studies, and figure-generation
scripts), configuration, and result CSVs that reproduce every number, table, and
figure in this paper are openly available at
\url{https://github.com/haodpsut/evokan-leo-opt}.

\bibliographystyle{IEEEtran}
\bibliography{references}

\vspace{-2mm}
\begin{IEEEbiographynophoto}{Phuc Hao Do}
received the Ph.D. degree in telecommunications. He is with the Faculty of
Information Technology, Da Nang Architecture University, Da Nang, Vietnam. His
research interests include networked AI, federated and quantum-inspired learning,
and signal processing for non-terrestrial networks.
\end{IEEEbiographynophoto}
\begin{IEEEbiographynophoto}{Truong Duy Dinh}
is with the Faculty of Information Security, Posts and Telecommunications Institute
of Technology, Ha Noi, Vietnam. His research interests include the Internet of
Things, network optimization, and machine learning for communications.
\end{IEEEbiographynophoto}

\end{document}
